---
title: "Homework 3"
format: pdf
editor: visual
author: "gabriela davidson"
date: "4-26-2026"
---

# Set up

```{r}
#| label: set_up
#| message: false

# read in tidyverse for data wrangling and visualization
library(tidyverse)
# read in here for reproducible file paths
library(here)
# read in janitor for cleaning column names
library(janitor)

# read in kelp data from data folder
kelp <- read_csv(here("data", "temp-kelp.csv"))

# read in personal data from data folder
my_data <- read_csv(here("data", "MyData_Homework3.csv"))
```

# Problem 1: Giant kelp fronds

## 1a. An appropriate test

The two appropriate tests to determine the strength of the relationship between ocean temperature and giant kelp frond elongation rate are the Pearson correlation coefficient and the Spearman rank correlation coefficient. The Pearson correlation measures the strength and direction of a linear relationship between two continuous variables and assumes that both variables are normally distributed. The Spearman rank correlation is a non-parametric alternative that measures the strength of a monotonic relationship between ocean temperature and elongation rate without assuming normality, making it appropriate if the normality assumption for Pearson is violated.

## 1b. Create a visualization

```{r}
#| label: kelp_visualization

# create scatterplot of ocean temperature vs kelp frond elongation rate
ggplot(data = kelp, # use kelp data frame
       aes(x = temp_c, # ocean temperature on x-axis
           y = kelp_elong)) + # elongation rate on y-axis
  geom_point(color = "darkgreen", # non-default point color
             size = 3,
             alpha = 0.7) + # slight transparency on points
  geom_smooth(method = "lm", # add linear trend line
              color = "darkorange", # non-default line color
              fill = "lightyellow", # confidence interval fill color
              se = TRUE) + # show confidence interval around line
  labs(x = "Ocean Temperature (°C)", # label x-axis with units
       y = "Kelp Frond Elongation Rate (cm day⁻¹)", # label y-axis with units
       title = "Relationship between ocean temperature and kelp frond elongation rate") +
  theme_bw() + # non-default theme
  theme(panel.grid.minor = element_blank()) # remove minor gridlines to reduce clutter
```

## 1c.Check your assumptions and run your test

```{r}

# check normality of ocean temperature using a QQ plot
ggplot(data = kelp,
       aes(sample = temp_c)) + # check temperature variable
  geom_qq() + # plot sample quantiles against theoretical quantiles
  geom_qq_line(color = "blue") + # add reference line for normality
  labs(title = "QQ plot: Ocean Temperature (°C)") +
  theme_bw() # non-default theme

# check normality of kelp elongation rate using a QQ plot
ggplot(data = kelp,
       aes(sample = kelp_elong)) + # check elongation rate variable
  geom_qq() + # plot sample quantiles against theoretical quantiles
  geom_qq_line(color = "blue") + # add reference line for normality
  labs(title = "QQ plot: Kelp Frond Elongation Rate (cm day⁻¹)") +
  theme_bw() # non-default theme

# run Pearson correlation test between ocean temperature and kelp elongation rate
cor.test(kelp$temp_c, # predictor variable: ocean temperature
         kelp$kelp_elong, # response variable: kelp frond elongation rate
         method = "pearson") # specify Pearson method for linear relationship
```

## 1d.

To evaluate the strength of the relationship between ocean temperature and giant kelp frond elongation rate, I used a Pearson correlation coefficient because both variables are continuous and approximately normally distributed, and the scatterplot suggested a linear relationship between them.
Ocean temperature was significantly negatively correlated with giant kelp frond elongation rate, meaning that kelp growing in warmer water tended to have slower growth rates (Pearson's r = -0.69, t(30) = -5.19, p < 0.001). This suggests that as ocean temperatures increase due to climate change, we would expect giant kelp fronds to grow more slowly, which could have important consequences for the overall health and structure of kelp forest ecosystems.

## 1e. Test implications
Our results indicate that higher ocean temperatures are associated with significantly lower giant kelp frond elongation rates, suggesting that ongoing ocean warming due to climate change may slow kelp growth. This is critical for our team to understand because reduced growth rates could diminish kelp forest biomass and structural complexity over time, threatening the many species that depend on these ecosystems. As ocean temperatures continue to rise, we should prioritize monitoring kelp growth rates in warmer regions as early indicators of broader kelp forest decline.

## 1f. Double check your own work
```{r}
#| label: kelp_spearman
# run Spearman rank correlation as the alternative non-parametric test
# does not assume normality, measures monotonic relationship
cor.test(kelp$temp_c, # predictor variable: ocean temperature
         kelp$kelp_elong, # response variable: kelp frond elongation rate
         method = "spearman") # specify Spearman method
```

Both the Pearson correlation (r = -0.69, t(30) = -5.19, p < 0.001) and the Spearman rank correlation (rho = -0.69, S = 9216.1, p < 0.001) led to the same decision to reject the null hypothesis that there is no relationship between ocean temperature and giant kelp frond elongation rate. Both tests revealed a strong negative relationship of nearly identical magnitude, meaning that both would lead to the same interpretation that higher ocean temperatures are associated with lower kelp frond elongation rates. The close agreement between the two tests strengthens confidence in the conclusion, as the Pearson result indicates the relationship is linear while the Spearman result confirms it is also monotonic.
